---
title: "Spotify Dataset Clustering - Profiling Analysis"
output: word_document
date: "2023-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Spotify Dataset Clustering - Profiling Analysis

This report is a continuation of the clustering.Rmd process. The SpotifyClustersData.csv was sourced from that dataset.

```{r}
# Replace with your own working directory if needed
WD <- "C:/Users/Joseph/Documents/Codes/2022/mvtec-2022/finalproject/04-clustering"
setwd(WD)

# dd <- read.csv("data/SpotifyClustersData.csv",header=T, sep=",", dec='.');

c2 <- read.csv("data/c2.csv", header=T) %>% pull()

dd <- read.csv("data/cleaneddata-withtitles.csv") %>%
  select("key", "tempo_cat",
         "loudness", "energy", "acousticness", "instrumentalness", "valence") %>%
  bind_cols(tibble(c2))
head(dd, 10)
```

As you may notice, the new cluster variable was generated from dissimilarity Matrix / Gower's Distance from previous script.

```{r}
names(dd)
```

```{r}
attach(dd)
```

Calculate the test value of variable Xnum for all modalities of factor P

```{r}
ValorTestXnum <- function(Xnum,P){
  # Freq dis of factors
  nk <- as.vector(table(P)); 
  n <- sum(nk);
  
  # Averages vs groups
  xk <- tapply(Xnum,P,mean);
  
  # Test values
  txk <- (xk-mean(Xnum))/(sd(Xnum)*sqrt((n-nk)/(n*nk)));
  
  # P-values
  pxk <- pt(txk,n-1,lower.tail=F);
  for(c in 1:length(levels(as.factor(P)))){if (pxk[c]>0.5){pxk[c]<-1-pxk[c]}}
  return (pxk)
}

ValorTestXquali <- function(P,Xquali){
  taula <- table(P,Xquali);
  n <- sum(taula); 
  pk <- apply(taula,1,sum)/n;
  pj <- apply(taula,2,sum)/n;
  pf <- taula/(n*pk);
  pjm <- matrix(data=pj,nrow=dim(pf)[1],ncol=dim(pf)[2], byrow=TRUE);      
  dpf <- pf - pjm; 
  dvt <- sqrt(((1-pk)/(n*pk))%*%t(pj*(1-pj)));
  
  # And there are divisions equal to 0 woman NA and it doesn't work
  zkj <- dpf
  zkj[dpf!=0]<-dpf[dpf!=0]/dvt[dpf!=0]; 
  pzkj <- pnorm(zkj,lower.tail=F);
  for(c in 1:length(levels(as.factor(P)))){for (s in 1:length(levels(Xquali))){if (pzkj[c,s]> 0.5){pzkj[c,s]<-1- pzkj[c,s]}}}
  return (list(rowpf=pf,vtest=zkj,pval=pzkj))
}
```

```{r}
# dades contain the dataset
dades <- dd

K<-dim(dades)[2]
par(ask=TRUE)
```

Using the added column 'cluster' which we'll continue use here as c2

```{r}
# P must contain the class variable
# c2 <- dd$cluster
P <- c2

nc <- length(levels(factor(P)))
nc
```

```{r}
pvalk <- matrix(data=0,nrow=nc,ncol=K, dimnames=list(levels(P),names(dades)))
nameP <- "Class"
n <- dim(dades)[1]
```

### Loop through the variable dataset to profile and visualize each

```{r}
for(k in 1:K){
  if (is.numeric(dades[,k])){ 
    print(paste("Analysis by classes of the Variable:", names(dades)[k]))
    
    boxplot(dades[,k]~P, main=paste("Boxplot of", names(dades)[k], "vs", nameP ), horizontal=TRUE)
    
    barplot(tapply(dades[[k]], P, mean),main=paste("Means of", names(dades)[k], "by", nameP ))
    abline(h=mean(dades[[k]]))
    legend(0,mean(dades[[k]]),"global mean",bty="n")
    print("Statistics per groups:")
    for(s in levels(as.factor(P))) {print(summary(dades[P==s,k]))}
    o<-oneway.test(dades[,k]~P)
    print(paste("p-value ANOVA:", o$p.value))
    kw<-kruskal.test(dades[,k]~P)
    print(paste("p-value Kruskal-Wallis:", kw$p.value))
    pvalk[,k]<-ValorTestXnum(dades[,k], P)
    print("p-values ValorsTest: ")
    print(pvalk[,k])      
  }else{
    if(class(dd[,k])=="Date"){
      print(summary(dd[,k]))
      print(sd(dd[,k]))
      
      # Decide breaks: weeks, months, quarters...
      hist(dd[,k],breaks="weeks")
      
    }else{
      # Qualitatives
      print(paste("Variable", names(dades)[k]))
      table<-table(P,dades[,k])

      rowperc<-prop.table(table,1)
      
      colperc<-prop.table(table,2)
      
      # Observe why the variable is true or false. It identifies the type of logic
      # This one has no levels, therefore, coercion was prevented
      dades[,k]<-as.factor(dades[,k])
      
      marg <- table(as.factor(P))/n
      print(append("Categories=",levels(as.factor(dades[,k]))))
      
      # From next plots, select one of them according to your practical case
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(colperc[,c],col=paleta[c]) }
      
      # With legend
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(colperc[,c],col=paleta[c]) }
      legend("topright", levels(dades[,k]), col=paleta, lty=2, cex=0.6)
      
      # Conditioned to classes
      print(append("Categories=",levels(dades[,k])))
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(rowperc[,c],col=paleta[c]) }
      
      # With legend
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]))
      paleta<-rainbow(length(levels(dades[,k])))
      for(c in 1:length(levels(dades[,k]))){lines(rowperc[,c],col=paleta[c]) }
      legend("topright", levels(dades[,k]), col=paleta, lty=2, cex=0.6)
      
      # With abcisses axis variable
      marg <-table(dades[,k])/n
      print(append("Categories=",levels(dades[,k])))
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]), las=3)
      paleta<-rainbow(length(levels(as.factor(P))))
      for(c in 1:length(levels(as.factor(P)))){lines(rowperc[c,],col=paleta[c]) }
      
      # With legend
      plot(marg,type="l",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]), las=3)
      for(c in 1:length(levels(as.factor(P)))){lines(rowperc[c,],col=paleta[c])}
      legend("topright", levels(as.factor(P)), col=paleta, lty=2, cex=0.6)
      
      # Conditioned to column
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]), las=3)
      paleta<-rainbow(length(levels(as.factor(P))))
      for(c in 1:length(levels(as.factor(P)))){lines(colperc[c,],col=paleta[c]) }
      
      # With legend
      plot(marg,type="n",ylim=c(0,1),main=paste("Prop. of pos & neg by",names(dades)[k]), las=3)
      for(c in 1:length(levels(as.factor(P)))){lines(colperc[c,],col=paleta[c])}
      legend("topright", levels(as.factor(P)), col=paleta, lty=2, cex=0.6)
      
      table<-table(dades[,k],P)
      print("Cross Table:")
      print(table)
      print("Conditional distributions column :")
      print(colperc)
      
      # Stacked bar charts                                        
      paleta<-rainbow(length(levels(dades[,k])))
      barplot(table(dades[,k], as.factor(P)), beside=FALSE,col=paleta )
      
      barplot(table(dades[,k], as.factor(P)), beside=FALSE,col=paleta )
      legend("topright",levels(as.factor(dades[,k])),pch=1,cex=0.5, col=paleta)
      
      # Attached bar charts
      barplot(table(dades[,k], as.factor(P)), beside=TRUE,col=paleta )
      
      barplot(table(dades[,k], as.factor(P)), beside=TRUE,col=paleta)
      legend("topright",levels(as.factor(dades[,k])),pch=1,cex=0.5, col=paleta)
      
      print("Square Chi test: ")
      print(chisq.test(dades[,k], as.factor(P)))
      
      print("Values test:")
      print( ValorTestXquali(P,dades[,k]))
      # Calculate the pvalues of quali
    }
  }
}
```

## Descriptions for every variable

### What patterns can we see for each class

-   Class 1 (black) - Most loudest and energetic

-   Class 2 (red) - Most acoustic

-   Class 3 (green) - Most energetic and instrumental

-   Class 4 (blue) - Most acoustic and instrumental

Happiest to saddest songs - Class 1 \> 2 \> 3 \> 4

```{r}
# Descriptors of the most significant classes. Add infoboxes
for (c in 1:length(levels(as.factor(P)))) {
  if(!is.na(levels(as.factor(P))[c])){
    print(paste("P.values per class:",levels(as.factor(P))[c]));
    print(sort(pvalk[c,]), digits=3) 
  }
}
```

Add the information of the modalities of qualitative to the list of pvalues and make global ordering

## Check song if fits profile

Based from the charts above, we describe the classes with these profiles:

- Class 1 - Mostly loudest

- Class 1 & 3 - Most energetic

- Class 2 & 4 - Most acoustic

- Class 3 & 4 - Most instrumental

- From Happiest to saddest Class 1, 2, 3, 4

We then picked the most popular song for every class and see if the profiling fits that song:

```{r}
dd_clust <- read.csv("data/cleaneddata-withtitles.csv") %>%
  bind_cols(tibble(cluster = c2))
head(dd_clust)
```

```{r}
names(dd_clust)
```
Here's the top 2 songs for every cluster. Try to listen to each song and check if it fits the profile we created.

```{r}
pop_per_class <- dd_clust %>%
  select(popularity, track_name, artists, track_genre, cluster) %>%
  arrange(desc(popularity)) %>%
  group_by(cluster) %>%
  top_n(6, popularity) %>%
  arrange(cluster)
pop_per_class
```
```{r}
ggplot(pop_per_class) +
  geom_bar(aes(x = cluster, fill = track_genre))
```
